\section{Conclusion}
\label{sec:conclusion}

We provide the first direct measurement of the Divergent Association Task
under controlled \layernorm interventions, across three pretrained Pre-LN
Transformers, three decoders, seven recipes, and four layer ranges. The
answer to the headline question --- ``does \layernorm cause creative
collapse?'' --- is partial and decoder-dependent. \layernorm does compress
concept-association geometry as \citet{brody2023expressivity} and
\citet{gupta2024geometric} predict, and removing its mean-subtraction does
expand concept-pair distance ($+7\%$ on \pythia). But the freed latent
variation only surfaces in \dat under high-temperature decoding, where it
lifts \gptsmall \dat by $+1.7$ to $+5.1$ points across all four layer
targets. Under greedy or moderate-$T$ top-$p$ sampling, no model shows a
consistent direction. After multiple-comparison correction, no individual
cell reaches significance.

\para{Key takeaway.}
\layernorm is one of several gates between latent variation and output
diversity, but not the dominant one. The LM-head and softmax recompress
most of the freed latent variation, and decoding sharpness gates whether
geometric room translates into behavioural diversity. Softening
\layernorm at inference is therefore not a free path to language-model
creativity gains.

\para{Future work.}
Five follow-ups directly address the limitations.
(i) \emph{From-scratch training.} Train tiny LMs (\eg 30M parameters on
TinyStories) with and without \layernorm and rerun \dat; this is the
cleanest causal test and was infeasible in our compute budget.
(ii) \emph{LM-head intervention.} Replace \layernorm ablation with an
orthogonal change-of-basis at the unembedding; if concept-pair expansion
\emph{can} be transmitted to output diversity, this experiment should
show it.
(iii) \emph{Larger, instruction-tuned models.} The concept-collapse story
is sharpest after RLHF; \textsc{Llama}-3-Instruct under \nomean with
min-$p$ sampling would test scale.
(iv) \emph{Algorithmic-creativity probes.} The Sibling/Triangle/Circle/Line
tasks of \citet{nagarajan2025roll} may be more sensitive to geometric
shifts than \dat, because they isolate compositional novelty separately
from word-level rarity.
(v) \emph{Joint LN--decoder ablation.} Holding \layernorm stock but
rotating the unembedding matrix would directly test whether the LM-head
is the dominant locus of collapse. We hypothesise it is.

The broader programme --- mapping the gates between latent variation and
output diversity --- is wide open. Our negative result on inference-time
\layernorm clears one candidate from the list and points downstream,
toward the LM head and the decoder.
