\section{Related Work}
\label{sec:related}

\para{Geometric role of \layernorm.}
\citet{brody2023expressivity} decompose
$\layernorm(\vx)=\gamma\odot(\vx-\mu)/\sigma+\beta$ into a projection (onto
the hyperplane orthogonal to $\bm{1}$) and a scaling onto a fixed-radius
sphere, and show that both components are needed for attention: the
projection enables uniform-attention queries (3$\times$ faster majority
computation), and the scaling fixes the \emph{unselectable-key} problem of
\citet{demeter2020stolen}. \citet{gupta2024geometric} extend this to seven
LLMs (including \pythia, \textsc{GPT-J}, and \textsc{Llama}) and show that
hidden states already have near-zero mean before \layernorm, so the mean
subtraction is empirically near-redundant --- which makes \rmsnorm
\citep{zhang2019root} a natural drop-in. \citet{menary2024transformer}
formalise the cross-coupling of semantic subspaces caused by Pre-Norm's
shared denominator, and report that O($10\%$) L2-norm perturbations
collapse $\sim\!1\%$ of attention heads. \citet{singhal2025impact} find
divergent roles in Pre-LN versus Post-LN: in Pre-LN, removing \layernorm
both destabilises learning and increases memorisation, with early-layer
\layernorm most influential. None of these studies measures the consequence
for creative-divergence metrics.

\para{Norm-removal architectures.}
A growing literature replaces or removes per-token normalisation:
\textsc{nGPT} \citep{loshchilov2025ngpt} forces every vector to unit norm;
\textsc{TaperNorm} \citep{kanavalau2026taper} introduces a gated
cosine-decay from \rmsnorm to a fixed linear map and identifies the
\emph{final} normalisation as the critical scale anchor; \textsc{DyT}
\citep{zhu2025dynamic} replaces \layernorm with dynamic element-wise
nonlinearities. These works study perplexity and downstream accuracy, not
divergent association. Our intervention recipes (\nomean, \noaffine,
\weakened, \identityln) are inference-time analogues of these training-time
designs.

\para{Creativity in LLMs.}
\citet{olson2021naming} introduce \dat as an objective measure of
associative creativity in humans, scoring the mean pairwise cosine distance
among $7$ of $10$ unrelated nouns via \glove. \citet{chen2023probing}
adapt \dat to LLMs, finding that \textsc{GPT-4} (greedy) scores $89.1$,
beating $96\%$ of $8{,}572$ human respondents.
\citet{bellemare2024divergent} confirm with $100{,}000$ humans across \dat
and creative writing. \citet{nagarajan2025roll} introduce
algorithmic-creativity tasks (Sibling/Triangle/Circle/Line) and report that
\emph{seed-conditioning} (input randomness) beats temperature for diversity
--- pointing to internal-state, not output-state, mechanisms. Our work asks
whether \layernorm is one such internal-state mechanism. \citet{he2025shakespearean}
probe creativity layer-wise via Layer-Skip and find an early-layer optimum
for the hallucination-creativity tradeoff. Our layer-targeted ablations
(\tgtearly, \tgtmid, \tgtlate) are the analogous lever at inference time.

\para{Concept geometry in latent space.}
The linear-representation hypothesis
\citep{marks2023geometry,park2024linear,geva2022transformer,belrose2023eliciting}
shows that concepts are encoded as linear directions in hidden states.
This work studies \emph{single} concepts; concept \emph{association}
(creative combination) requires interaction geometry. Our per-layer
pairwise-cosine probe over a noun bank is the natural extension.

\para{Decoding for diversity.}
\citet{holtzman2020curious} establish nucleus (top-$p$) sampling;
\citet{nguyen2025minp} introduce min-$p$ sampling. These act on the output
distribution. Our finding that \layernorm's geometric expansion only
propagates to \dat at high temperature is consistent with the
\emph{joint-gating} story: latent room must coexist with a decoder loose
enough to surface it.

\para{Positioning.}
Geometric work establishes \emph{what} \layernorm does; behavioural work
establishes \emph{how to measure} associative creativity. We cross the
streams: controlled \layernorm interventions on pretrained Pre-LN models,
with \dat, perplexity, and per-layer geometry measured jointly. To our
knowledge this is the first such study.
